---
title: "SPD ALPR Analysis"
author:
- '[Destiny Moreno](https://github.com/dmorenouw)'
- '[Phil Neff](https://github.com/philneff)'
date: "20 April, 2023"
output:
  html_document:
    html_preview: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: no
    toc_depth: '3'
---

```{r import, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}

library(pacman)
p_load(tidyverse) 
p_load(lubridate) 
p_load(ggmap) 
p_load(kableExtra)
p_load(ggrepel)
p_load(scales)
p_load(leaflet)
p_load(leaflet.extras)
p_load(tigris)
p_load(sf)
p_load(viridis)


options(scipen = 999)

df <- read_delim(here::here('write', 'input', 'spd-alpr-pub.csv.gz'), delim='|',
                    col_types = cols(dt = col_datetime(format="%m/%d/%Y - %H:%M:%S")
                                  ))

df <- df %>% 
  mutate(year = year(dt),
         month = month(dt),
         day = day(dt),
         hour = hour(dt),
         minute = minute(dt),
         wday = wday(dt),
         date = ymd(paste(year, month, day, sep='-')),
         day_name = lubridate::wday(date, label = TRUE, abbr = FALSE))

ampm <- function(hr) paste((hr-1) %% 12 + 1, ifelse(hr %% 24 < 12, "am", "pm"))

df$missing_address = is.na(df$address)

labels <- read_delim(here::here('write', 'input', 'labels.csv'))

ggkey = Sys.getenv("GEOCODE_API_KEY")
register_google(key = ggkey)


```

# Introduction 

This notebook analyzes one week of Seattle Police Department (SPD) Automated License Plate Reader (ALPR) activity obtained by the University of Washington Center for Human Rights (UWCHR) through public records requests.

# License Plate Reads

Each time a license plate is detected by the ALPR device, the system records the date and time of the detection and the location of the license plate hit. This information reveals insights about how these surveillance devices are used.

```{r descriptive stats:reads, echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}

total_reads  <- nrow(df)

dtrange <- range(df$dt)

unique_plates <- length(unique(df$plate_hash, na.rm=TRUE))
unique_plates_p <- unique_plates / total_reads *100

plate_freq <- df %>% 
  group_by(plate_hash) %>% 
  summarize(n = n()) %>% 
  mutate(freq = n/sum(n))

avg_reads <- mean(plate_freq$n)

one_read <- plate_freq %>% 
  filter(n == 1) %>% 
  nrow()
one_read_p <- one_read / unique_plates *100

multi_read <- plate_freq %>% 
  filter(n > 1) %>% 
  nrow()
multi_read_p <- multi_read / unique_plates *100

date_1 <- paste0(month(dtrange[1], label = TRUE, abbr = FALSE), ' ', day(dtrange[1]), ', ', year(dtrange[1]))

date_2 <- paste0(month(dtrange[2], label = TRUE, abbr = FALSE), ' ', day(dtrange[2]), ', ', year(dtrange[2]))

```

The dataset contains **`r total_reads`** instances where potential license plates were detected by SPD's surveillance system from **`r date_1`** (a `r wday(dtrange[1], label=TRUE, abbr=FALSE)`) to **`r date_2`** (a `r wday(dtrange[2], label=TRUE, abbr=FALSE)`). Of the license plates captured, **`r one_read`** reads, which accounts for **`r round(one_read_p, 0)`%** of the total reads, were only captured once. We also found that on average, each unique license plate was captured **`r round(avg_reads, 2)`** times.

ALPR cameras also appear to pick up other types of visual data. For instance, the data includes reads that appear like road signs (such as *AHEAD, DETOUR, CLOSED, ROAD*, and *VVARNING*) and other public signage (like *POLICE, TOWAWAY, CROSSWALK*, and *PARKING*). Though these reads did not capture license plates, we did not remove them from our analysis because they still provide meaningful information about the presence of ALPR readers. 

# Analyzing the Spatial Deployment of SPD's ALPR System

**Figure 1** provides an overview of the geographic distribution of these reads. As a note, reads are based on the geolocation of each address provided by SPD, which may not be exact. 

One ALPR device is observed to travel outside of Seattle and King County in the vicinity of the cities of Everett and Snohomish.

### Figure 1: Overview of All Reads

```{r leaflet map all reads,  echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}

plot_color = "#4b2e83"

#Build  map 
leaflet_all_reads <- leaflet(df) %>% 
  addProviderTiles(providers$CartoDB.VoyagerLabelsUnder) %>%
  addCircleMarkers(
    radius = 2,
    color = "black",
    fillColor = plot_color,
    stroke = TRUE,  weight = .3,
    fillOpacity = 0.1) %>%
  addMarkers(
    lng = -122.3345865, lat = 47.70302542,
    label = "Seattle Police Department, North Precinct",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "sans serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) %>%
  addMarkers(
    lng = -122.299097, lat = 47.66211867,
    label = "University Village",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "san serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) %>%
  addMarkers(
    lng = -122.3038448, lat = 47.53708176,
    label = "King County Airport",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "sans serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) %>%
  addMarkers(
    lng = -122.3159617, lat = 47.67653678,
    label = "Roosevelt Light Rail Station",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "sans serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) %>%
    addMarkers(
    lng = -122.3364106, lat = 47.61620391,
    label = "Seattle Police Department, West Precinct",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "sans serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)")))
  # %>%
  # addMarkers(
  #   lng = -121.9996418, lat = 47.84514158,
  #   label = "Monroe Correctional Complex",
  #   labelOptions = labelOptions(noHide = T, direction = "bottom",
  #     style = list(
  #       "color" = "black",
  #       "font-family" = "sans serif",
  #       "font-style" = "bold",
  #       "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
  #       "font-size" = "12px",
  #       "border-color" = "rgba(0,0,0,0.5)"))) 

leaflet_all_reads

```

For a closer look, **Figure 2**, **Figure 3**, and **Figure 4** depict zoomed-in maps that highlight areas with the highest concentration of ALPR reads through brighter colors. The most read-dense areas are located near SPD precincts. Notably, there are no reads in West Seattle and very few reads in East Capitol Hill and waterfront neighborhoods like Magnolia. This may be due to missing data, as discussed below. 

### Figure 2: Heat Map of ALPR Reads

License plates are more likely to be captured on camera in the vicinity of SPD precincts.

```{r heatmap all reads,  echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}

# Build base map of Seattle
seattle <- ggmap(get_googlemap(center = c(lon = -122.332115, 
           lat = 47.636462),zoom = 11, scale = 1, 
           maptype ='roadmap')
           ) 

# Plot all observations onto Seattle map
heatmap_all_reads <- seattle + 
  stat_density2d(aes(x = lon, y = lat, 
                     fill = ..level.., alpha = ..level..), 
                 bins = 100, data = df, geom = "polygon") + 
  scale_fill_distiller(palette = "Spectral") +
  theme_void() + 
  theme(legend.position="none",
             plot.caption = element_text(face = "italic")) + 
  labs(caption = "Source: SPD")  

heatmap_all_reads

```

### Figure 3: Heat Map of ALPR Reads in Downtown Seattle

While license plates all throughout downtown are likely to be captured by the ALPR system, we have observed that specific neighborhoods, such as East Capitol Hill, appear to go entirely un-monitored, at least during the time period covered by this dataset.

```{r downtown heat map, echo=FALSE, message=FALSE, warning=FALSE}

heatmap_downtown <- ggmap(
  get_googlemap(center = c(lon = -122.334480, lat = 47.605026), 
                zoom = 15, scale = 1, maptype ='roadmap', color = 'color')) +
  stat_density2d(
    aes(x = lon, y = lat, fill = ..level.., alpha = ..level..),
    data = df,geom = "polygon") + 
  scale_fill_distiller(palette = "Spectral") +
   theme_void()  +
  theme(legend.position="none",
             plot.caption = element_text(face = "italic")) + 
    labs(caption = "Source: SPD") 
  
heatmap_downtown

```

### Figure 4: Heat Map of ALPR Reads near SPD Precinct

SPD's North Precinct, marked by the white point in the map below, is located near the Northgate Transit Center. The higher frequency of reads in this area could be attributed to patrol cars traveling to and from the precinct.

```{r spd north heat map, echo=FALSE, message=FALSE, warning=FALSE}

heatmap_spdnorth <- ggmap(
  get_googlemap(
    center = c(lon = -122.334724, lat = 47.702875), 
    zoom = 15, scale = 1, maptype ='roadmap', color = 'color')) + 
  stat_density2d(
    aes(x = lon, y = lat, fill = ..level.., alpha = ..level..),
    bins = 1000, data = df,
    geom = "polygon") + 
  scale_fill_distiller(palette = "Spectral") +
  geom_point(
    aes(x = lon, y = lat), data = labels, color = "black", fill="white",
    shape= 23, size = 2, stroke = 1) +
  theme_void()  +
  theme(
    legend.position="none",
    plot.caption = element_text(face = "italic")) + 
  labs(caption = "Source: SPD") 

heatmap_spdnorth

```

### Figure 5: Heat Map of ALPR Reads near University Village

High frequency of APLR reads in the vicinity of the University of Washington Seattle campus and University Village appear to reflect a device capturing plate reads from a stationary position near NE 45th Street.

```{r university village, echo=FALSE, message=FALSE, warning=FALSE}

heatmap_uvillage <- ggmap(
  get_googlemap(center = c(lon = -122.300089, lat = 47.661616), 
                zoom = 15, scale = 1, maptype ='roadmap', color = 'color')) +
  stat_density2d(
    aes(x = lon, y = lat, fill = ..level.., alpha = ..level..),
    bins = 100,
    data = df,geom = "polygon") + 
  scale_fill_distiller(palette = "Spectral") +
   theme_void()  +
  theme(legend.position="none",
             plot.caption = element_text(face = "italic")) + 
    labs(caption = "Source: SPD") 

heatmap_uvillage

```

## ALPR Device Characteristics

```{r describe device and reads, echo=FALSE, message=FALSE, include=FALSE}

#Label devices as they appear in the data (called names using unique function and labeled in the same order)
df <- df %>%
  mutate(device_abbr = factor(device) %>%
           fct_recode("A" = "M207541_Deleted_2021-10-21_22 50:42",
                      "B" = "SPD2020681MDC",
                      "C" = "BXH9904 SPD2020627MDC",
                      "D" = "M207709",
                      "E" = "SPD2020537MDC",
                      "F" = "SPD2020602MDC",
                      "G" = "PEOATG508533",
                      "H" = "SPD2020621MDC",
                      "I" = "PEO508459" ))

reads_per_device<- df %>%
  group_by(device, device_abbr) %>%
  summarise(n=n(), 
            missing_address=sum(missing_address)) %>%
  mutate(percent = n/nrow(df) *100) %>%
  arrange(desc(percent))

#Distribution of hits per zip code per device
reads_per_device_per_zip <- df %>%
  group_by(device_abbr, zipcode) %>%
  summarise(n=n()) %>%
  mutate(percent = n/nrow(df) *100) %>%
  arrange(desc(percent))

max_reads_per_device <- max(reads_per_device$n)

avg_reads_per_device <- mean(reads_per_device$n)

device_count <- length(unique(reads_per_device$device_abbr))

reads_missing_address <- df %>%
  filter(missing_address %in% c('TRUE'))

devices_missing_address = reads_missing_address %>%
  filter(device_abbr== "H" | device_abbr == "I")  %>%
  nrow()
devices_missing_address_p = devices_missing_address / total_reads * 100

peo_device_reads = sum(df$device_abbr== "G" | df$device_abbr == "F") 
peo_device_reads_p = peo_device_reads / total_reads * 100


```

The data set consists of **`r device_count`** devices. Each device captured an average of **`r round(avg_reads_per_device, 0)`** license plates, with the most device capturing **`r max_reads_per_device`** plates.

Reads associated with two devices were completely missing addresses, comprising **`r round(devices_missing_address_p, 0)`%** of all reads in the data.

Another two devices were attached to parking enforcement officer vehicles (PEO). Reads from PEO devices account for **`r round(peo_device_reads_p, 0)`%** of reads in the data set.

### Figure 6: Reads Per Device

```{r table of devices, echo=FALSE, message=FALSE}

table_reads_per_device <- kable(reads_per_device, 
                                col.names=c('Device', "Abbr.", 'Reads', 
                                            'Missing Address', 'Percent of Reads'),
                                row.names = F) %>%
     kableExtra::kable_styling(
       bootstrap_options = "bordered",
       position = "center",
       full_width = FALSE
     )

table_reads_per_device

```

### Figure 7: Tracking the Movements of ALPR Devices

The path of two active ALPR devices are depicted below. The maps show where the device captured license plates on camera througout the day. This vehicle collected plates all throughout the Seattle area across multiple days.

```{r map device path, echo=FALSE, message=FALSE, warning=FALSE}

# Create base of Seattle
base <- get_map("seattle, washington", zoom = 11, maptype = c("roadmap"))

# Filter data for specific device
device_M207541 <- df %>%
  arrange(device_abbr, dt) %>%
  filter(device_abbr == "A")

# Map device path 
date_label <- c(
  '2021-10-01'="Friday",
  '2021-10-02'="Saturday",
  '2021-10-03'="Sunday",
  '2021-10-04'="Monday",
  '2021-10-05'="Tuesday",
  '2021-10-06'="Wednesday",
  '2021-10-07'="Thursday",
  '2021-10-08'="Friday"
)

device_M207541_map <- ggmap(base) + 
  geom_path(data = device_M207541, aes(x = lon, y = lat, color = hour(dt),
                                       group = device_abbr)) + 
  facet_wrap(~date, strip.position = "bottom",
             labeller = as_labeller(date_label)) +
  theme_void() + 
   theme(
     plot.subtitle = element_text(face = "bold"),      
     plot.caption = element_text(face = "italic"), 
     legend.key.width = unit(30, "pt"), 
     legend.position	= "bottom", 
     legend.justification	= "center", 
     legend.box = "horizontal",
     legend.margin = margin(6, 6, 6, 6), 
     strip.background = element_rect(colour = "black", 
                                     fill = "white", size = .2),
     strip.text.x = element_text(face = "bold", size = 10, 
                                margin = margin(4, 2, 4, 2, "pt")),
     panel.spacing = unit(10, "pt")) + 
    labs(
         subtitle = "Device 'D'",
         caption = "Source: SPD",
         color = "Time of Day")

device_M207541_map

```

Meanwhile, this vehicle was active during just one day in the data set, and it only captured plates during a trip outside Seattle and King County.

```{r map device outside seattle, echo=FALSE, message=FALSE, warning=FALSE}

everett <- get_map("everett, washington", zoom = 11, maptype = c("roadmap"))

device_SPD2020621MDC <- df %>%
  arrange(device_abbr, dt) %>%
  filter(device_abbr == "H")

device_SPD2020621MDC_map <- ggmap(everett) + 
  geom_path(data = device_SPD2020621MDC, aes(x = lon, y = lat, color = hour(dt),
                                       group = device_abbr)) + 
  facet_wrap(~date, strip.position = "bottom",
             labeller = as_labeller(date_label)) +
  theme_void() + 
   theme(
     plot.subtitle = element_text(face = "bold"),      
     plot.caption = element_text(face = "italic"), 
     legend.key.width = unit(30, "pt"), 
     legend.position	= "bottom", 
     legend.justification	= "center", 
     legend.box = "horizontal",
     legend.margin = margin(6, 6, 6, 6), 
     strip.background = element_rect(colour = "black", 
                                     fill = "white", size = .2),
     strip.text.x = element_text(face = "bold", size = 10, 
                                margin = margin(4, 2, 4, 2, "pt")),
     panel.spacing = unit(10, "pt")) + 
    labs(
      subtitle = "Device 'A'",
      caption = "Source: SPD",
      color = "Time of Day")

device_SPD2020621MDC_map

```

# ALPR System Usage By Day and Time

```{r describe date and time, echo=FALSE, message=FALSE}

reads_per_day <- df %>% 
  group_by(wday) %>% 
  summarize(n = n()) 
max_reads_per_day <- max(reads_per_day)

reads_per_min <- df %>% 
  group_by(day, hour, minute) %>% 
  summarize(n = n()) 
max_reads_per_min <- max(reads_per_min)

```

Contrary to our expectations, SPD devices were overall relatively inactive during the beginning of the work day and inactive during the morning rush hour commute. **Figure 8** and **Figure 9** show that devices collected the most data in the middle of the work day, between **12:00 pm** and **6:00 pm**. On the busiest day in our data set, SPD cameras captured **`r max_reads_per_day`** license plates. Up to **`r max_reads_per_min`** plates were captured in a single minute. 

### Figure 8: SPD ALPR Collection by Time of Day

ALPR activity was low in the early morning and evening. Overall, ALPR activity was busiest in the middle of the day. 

```{r bar plot reads by hour, echo=FALSE, message=FALSE, warning=FALSE}

bar_reads_per_hour <- df %>%
  ggplot( aes(x=hour)) +
    geom_bar(bins = 24, binwidth=1, 
                    fill="#4b2e83", color="#4b2e83", alpha=0.9) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = ampm,
    breaks = function(z) pretty(z, n = 12)) +
         theme(
            axis.title = element_text(face = "bold"),
            axis.text.x = element_text(margin = margin(t = .3, unit = "cm")),
           plot.caption = element_text(face = "italic")) +
   labs(y = "Reads per Hour", 
        x = "Hour of Day",
        caption = "Source: SPD")

bar_reads_per_hour

``` 

### Figure 9: SPD ALPR Collection by Time of Day Per Day

This trend was similar throughout the week. 

```{r bar plot reads by hour per day, echo=FALSE, message=FALSE, warning=FALSE}

bar_reads_per_hour_per_day <- df %>%
  ggplot( aes(x=hour)) +
    geom_bar(bins = 24, binwidth=1, 
                    fill="#4b2e83", color="#4b2e83", alpha=0.9) +
    facet_wrap(~date,  
               ncol = 4,
               shrink = FALSE,
               strip.position = "bottom",
               labeller = as_labeller(date_label)) + 
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = ampm,
                     breaks = function(z) pretty(z, n = 12)) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 90), 
    strip.background = element_rect(colour = "black",fill = "white", size = .2),
    strip.text.x = element_text(face = "bold", size = 8, 
                                margin = margin(4, 2, 4, 2, "pt")),
    plot.caption = element_text(face = "italic")) +
   labs(y = "Reads per Hour", 
        x = "Day of the Week",
        caption = "Source: SPD")

bar_reads_per_hour_per_day

```

### Figure 10: SPD ALPR Collection by Day of the Week

ALPR activity was busier on weekdays, with a lower frequency of ALPR captures occurring over the weekend. 

```{r bar plot reads by day, echo=FALSE, message=FALSE, warning=FALSE}

bar_reads_per_day <- df %>%
  ggplot(aes(x=wday)) +
    geom_bar( bins = 7, binwidth=.5, 
                    fill="#4b2e83", color="#4b2e83", alpha=0.9) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(1,7, by = 1), labels = c(
    "Sunday","Monday","Tuesday","Wednesday","Thursday", "Friday", "Saturday")) +
  theme(
            axis.title = element_text(face = "bold"),
            axis.text.x = element_text(margin = margin(t = .3, unit = "cm")),
           plot.caption = element_text(face = "italic")) +
  labs(y = "Reads Per Day", 
          x = "Day of the Week",
          caption = "Source: SPD") 

bar_reads_per_day

```

### Figure 11: SPD ALPR Collection by Day and Device

Not all ALPR devices recorded license plate reads every day of the week. Note devices "C" and "E" are missing all address data.

```{r bar plot reads by day and device, echo=FALSE, message=FALSE, warning=FALSE}

# Proportions of the bar colored by devices

bar_reads_per_day_device <- df %>%
  ggplot(aes(x=wday)) +
  geom_bar(aes(fill = device_abbr),
           bins = 7, binwidth = .5) +
    scale_y_continuous(labels = comma) +
    scale_x_continuous(breaks = seq(1,7, by = 1), labels = c(
      "Sunday","Monday","Tuesday","Wednesday","Thursday", "Friday", "Saturday")) +
  scale_fill_viridis_d() +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(margin = margin(t = .3, unit = "cm")),
    plot.caption = element_text(face = "italic")) +
     labs(y = "Reads Per Day", 
        x = "Day of the Week",
        caption = "Source: SPD")

plotly::ggplotly(bar_reads_per_day_device)

```


# Analyzing SPD's ALPR System and Census Data 

Our preliminary statistical analysis suggests a weak negative correlation between plate hits and median household income, indicating a possible increase of surveillance in lower income neighborhoods. **Figure 12** below depicts this relationship; the darker the blue, the higher the intensity of ALPR reads. We did not identify any other suggestive correlations between SPD surveillance and other demographic factors.

However, when accounting for over-dispersion, the correlation loses its statistical significance. This finding should thus be interpreted with caution. Moreover, our analysis only covers data from a single week, and a large portion of the collected readings are missing important location data. For a brief analysis, see this [companion notebook](analyze.html).

```{r map of median income, message=FALSE, warning=FALSE, include=FALSE}

# Retrieve physical attributes for map from tigris package
my_state = 'WA'
my_county = c('King', 'Snohomish','Kitsap','Pierce','Island')
my_roads = primary_secondary_roads(my_state) %>% 
               filter(RTTYP %in% c('U','S','I','C','O'))
landmass = counties(cb = TRUE, state = my_state)
water = area_water(my_state,my_county)

#Create basic map from physical attributes
my_map = ggplot() +
  geom_sf(data = landmass, fill = '#e6e6e6', color = NA, 
          aes(geometry = geometry)) + 
  geom_sf(data = water, fill = '#cce6ff', color = NA, 
          aes(geometry = geometry)) + 
  geom_sf(data = my_roads, color = 'white', aes(geometry = geometry)) +
  theme(
        plot.title = element_text(family = "Helvetica",
                                  face = "bold",
                                  colour = "black",
                                  size = 12,
                                  margin = margin(1),
                                  hjust = .5),
        plot.subtitle = element_text(family = "Helvetica",
                                  colour = "black",
                                  size = 9,
                                  margin = margin(1),
                                  hjust = .5),
        plot.caption = element_text(family = "Helvetica",
                                  colour = "black",
                                  size = 6,
                                  margin = margin(1),
                                  hjust = .5),
        # legend.justification=c(1,0), legend.position=c(1,0),
        legend.title = element_text(family = "Helvetica",
                                  colour = "black",
                                  size = 6,
                                  margin = margin(1),
                                  hjust = .5),
        legend.text = element_text(family = "Helvetica",
                                  colour = "black",
                                  margin = margin(1),
                                  size = 4,
                                  hjust = .5),
        legend.spacing.y = unit(0, "mm"),
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "black"),
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

# Merge geography data for map (zip tabulations = zip codes)
shapefile <- zctas(cb = F, year = 2010, state = 'WA') %>%
  subset(select = c(ZCTA5CE10, geometry)) %>%
  mutate(zipcode = ZCTA5CE10)

df <- merge(df, shapefile, by = "zipcode", all.x = TRUE) 

limits = st_bbox(df$geometry)

boundaries_df <- coord_sf(xlim = c(as.numeric(limits$xmin),as.numeric(limits$xmax)), 
           ylim = c(as.numeric(limits$ymin),as.numeric(limits$ymax)))


#Create choropleth map 
medincome_map = my_map +
  scale_fill_distiller(direction = -1) +
  geom_sf(data = df %>%
            filter(!is.na(median_incomeE)), 
          color = NA, aes(geometry = geometry, fill = median_incomeE)) 

#Plot ALPR reads onto choropleth map: df 
medincome_boundaries_df <- medincome_map + boundaries_df 

medincome_p1 <- medincome_boundaries_df + 
  geom_point(aes(x = lon, y = lat), 
             data = df,
             # filter(is.na(outlier)),
             color = "black", fill= plot_color, shape= 21, size = .5, stroke = .03) + 
  labs(
       subtitle = "",
       fill = "2021 Median Household Income",
       caption = "Source: [SPD] and U.S. Census Bureau 2021 Household Income Estimate") 
```

### Figure 12: ALPR Plate Reads and Median Income

```{r calling map of median income, echo=FALSE, message=FALSE, warning=FALSE}
medincome_p1

```


