---
title: "SPD ALPR Analysis"
author:
- '[Destiny Moreno](https://github.com/dmorenouw)'
- '[Phil Neff](https://github.com/philneff)'
date: "20 April, 2023"
output:
  html_document:
    html_preview: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: no
    toc_depth: '3'
---

```{r import, echo=FALSE, message=FALSE, include=TRUE}

library(pacman)
p_load(tidyverse) 
p_load(lubridate) 
p_load(ggmap) 
p_load(kableExtra)
p_load(ggrepel)
p_load(scales)
p_load(leaflet)
p_load(leaflet.extras)

options(scipen = 999)

df <- read_delim(here::here('write', 'input', 'spd-alpr-pub.csv.gz'), delim='|',
                    col_types = cols(dt = col_datetime(format="%m/%d/%Y - %H:%M:%S")
                                  ))

df <- df %>% 
  mutate(year = year(dt),
         month = month(dt),
         day = day(dt),
         hour = hour(dt),
         minute = minute(dt),
         wday = wday(dt),
         date = ymd(paste(year, month, day, sep='-')),
         day_name = lubridate::wday(date, label = TRUE, abbr = FALSE))

ampm <- function(hr) paste((hr-1) %% 12 + 1, ifelse(hr %% 24 < 12, "am", "pm"))

df$missing_address = is.na(df$address)

labels <- read_delim(here::here('write', 'input', 'labels.csv'))

ggkey = Sys.getenv("GEOCODE_API_KEY")
register_google(key = ggkey)


```

# Introduction 

The data set captures one week of Seattle Police Department ALPR activity obtained through public records requests.

# Variables

## License Plate Reads

Each time a license plate is detected by the ALPR device, the system records the date and time of the detection and the location of the license plate hit, including the exact longitude and latitude coordinates and the full address. This information revealed insight about the frequency and intensity of the surveillance program and the areas where it is used.


```{r descriptive stats:reads, echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}

total_reads  <- nrow(df)

dtrange <- range(df$dt)

unique_plates <- length(unique(df$plate_hash, na.rm=TRUE))
unique_plates_p <- unique_plates / total_reads *100

plate_freq <- df %>% 
  group_by(plate_hash) %>% 
  summarize(n = n()) %>% 
  mutate(freq = n/sum(n))

avg_reads <- mean(plate_freq$n)

one_read <- plate_freq %>% 
  filter(n == 1) %>% 
  nrow()
one_read_p <- one_read / unique_plates *100

multi_read <- plate_freq %>% 
  filter(n > 1) %>% 
  nrow()
multi_read_p <- multi_read / unique_plates *100

```

# Analysis 

Our data set contains **`r total_reads`** instances where license plates were detected by SPD's surveillance system from **`r dtrange[1]`** (a Friday) to **`r dtrange[2]`** (a Friday). Of the license plates captured, **`r one_read`** reads, which accounts for **`r round(one_read_p, 0)`%** of the total reads, were only captured once. We also found that on average, each unique license plate was captured **`r round(avg_reads, 2)`** times.

ALPR cameras also appear to pick up other types of visual data. For instance, the data includes reads that appear like road signs (such as *AHEAD, DETOUR, CLOSED, ROAD*, and *VVARNING*) and other public signage (like *POLICE, TOWAWAY, CROSSWALK, *and *PARKING*). Though these reads did not capture license plates, we did not remove them from our analysis because they still provide meaningful information about the presence of ALPR readers. 

## Analyzing the Spatial Deployment of SPD's ALPR System

**Figure 1** provides an overview of the geographic distribution of these reads. As a note, reads are based on the geolocation of each address provided by SPD, which may not be exact. 

Nearly all of the non-Seattle reads are located along the journey to a prison north of Seattle, the Monroe Correctional Complex. 


### Figure 1: Overview of All Reads

```{r leaflet map all reads,  echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}

plot_color = "#4b2e83"

#Build  map 
leaflet_all_reads <- leaflet(df) %>% 
  addProviderTiles(providers$CartoDB.VoyagerLabelsUnder) %>%
  addCircleMarkers(
    radius = 2,
    color = "black",
    fillColor = plot_color,
    stroke = TRUE,  weight = .3,
    fillOpacity = 0.1) %>%
  addMarkers(
    lng = -122.3345865, lat = 47.70302542,
    label = "Seattle Police Department, North Precinct",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "sans serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) %>%
  addMarkers(
    lng = -122.299097, lat = 47.66211867,
    label = "University Village",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "san serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) %>%
  addMarkers(
    lng = -122.3038448, lat = 47.53708176,
    label = "King County Airport",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "sans serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) %>%
  addMarkers(
    lng = -122.3159617, lat = 47.67653678,
    label = "Roosevelt Light Rail Station",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "sans serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) %>%
    addMarkers(
    lng = -122.3364106, lat = 47.61620391,
    label = "Seattle Police Department, West Precinct",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "sans serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) %>%
  addMarkers(
    lng = -121.9996418, lat = 47.84514158,
    label = "Monroe Correctional Complex",
    labelOptions = labelOptions(noHide = T, direction = "bottom",
      style = list(
        "color" = "black",
        "font-family" = "sans serif",
        "font-style" = "bold",
        "box-shadow" = "3px 3px rgba(0,0,0,0.25)",
        "font-size" = "12px",
        "border-color" = "rgba(0,0,0,0.5)"))) 

leaflet_all_reads

```

For a closer look, **Figure 2**, **Figure 3**, and **Figure 4** depict zoomed-in maps that highlights areas with the highest concentration of ALPR reads through brighter colors. The most read-dense areas are located near SPD precincts. Notably, there are no reads in West Seattle and very few reads in East Capitol Hill and waterfront neighborhoods like Magnolia. This may be due to missing data. 

### Figure 2: Heat Map of ALPR Reads

License plates are more likely to be captured on camera in the vicinity of SPD precincts.

```{r heatmap all reads,  echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}

# Build base map of Seattle
seattle <- ggmap(get_googlemap(center = c(lon = -122.332115, 
           lat = 47.636462),zoom = 11, scale = 1, 
           maptype ='roadmap')
           ) 

# Plot all observations onto Seattle map
heatmap_all_reads <- seattle + 
  stat_density2d(aes(x = lon, y = lat, 
                     fill = ..level.., alpha = ..level..), 
                 bins = 100, data = df, geom = "polygon") + 
  scale_fill_distiller(palette = "Spectral") +
  theme_void() + 
  theme(legend.position="none",
             plot.caption = element_text(face = "italic")) + 
  labs(caption = "Source: SPD")  

heatmap_all_reads

```

### Figure 3: Heat Map of ALPR Reads in Downtown Seattle

While license plates all throughout downtown are likely to be captured by the ALPR system, we have observed that specific neighborhoods, such as East Capitol Hill, appear to go entirely unmonitored.

```{r downtown heat map, echo=FALSE, message=FALSE, warning=FALSE}

heatmap_downtown <- ggmap(
  get_googlemap(center = c(lon = -122.334480, lat = 47.605026), 
                zoom = 15, scale = 1, maptype ='roadmap', color = 'color')) +
  stat_density2d(
    aes(x = lon, y = lat, fill = ..level.., alpha = ..level..),
    data = df,geom = "polygon") + 
  scale_fill_distiller(palette = "Spectral") +
   theme_void()  +
  theme(legend.position="none",
             plot.caption = element_text(face = "italic")) + 
    labs(caption = "Source: SPD") 
  
heatmap_downtown

```

### Figure 4: Heat Map of ALPR Reads by SPD Precint

SPD's North Precinct, marked by the white point in the map below, is located near the Northgate Transit Center. The higher frequency of reads in this area could be attributed to patrol cars traveling to and from the precinct.

```{r spd north heat map, echo=FALSE, message=FALSE, warning=FALSE}

heatmap_spdnorth <- ggmap(
  get_googlemap(
    center = c(lon = -122.334724, lat = 47.702875), 
    zoom = 15, scale = 1, maptype ='roadmap', color = 'color')) + 
  stat_density2d(
    aes(x = lon, y = lat, fill = ..level.., alpha = ..level..),
    bins = 1000, data = df,
    geom = "polygon") + 
  scale_fill_distiller(palette = "Spectral") +
  geom_point(
    aes(x = lon, y = lat), data = labels, color = "black", fill="white",
    shape= 23, size = 2, stroke = 1) +
  theme_void()  +
  theme(
    legend.position="none",
    plot.caption = element_text(face = "italic")) + 
  labs(caption = "Source: SPD") 

heatmap_spdnorth

```

### Figure 5: Heat Map of ALPR Reads by University Village

```{r university village, echo=FALSE, message=FALSE, warning=FALSE}

heatmap_uvillage <- ggmap(
  get_googlemap(center = c(lon = -122.300089, lat = 47.661616), 
                zoom = 15, scale = 1, maptype ='roadmap', color = 'color')) +
  stat_density2d(
    aes(x = lon, y = lat, fill = ..level.., alpha = ..level..),
    bins = 100,
    data = df,geom = "polygon") + 
  scale_fill_distiller(palette = "Spectral") +
   theme_void()  +
  theme(legend.position="none",
             plot.caption = element_text(face = "italic")) + 
    labs(caption = "Source: SPD") 

heatmap_uvillage

```
## Investigating the ALPR Devices in Use

```{r describe device and reads, echo=FALSE, message=FALSE, include=FALSE}

reads_per_device<- df %>%
  group_by(device) %>%
  summarise(n=n()) %>%
  mutate(percent = n/nrow(df) *100) %>%
  arrange(desc(percent))

#Distribution of hits per zip code per device
reads_per_device_per_zip <- df %>%
  group_by(device, zipcode) %>%
  summarise(n=n()) %>%
  mutate(percent = n/nrow(df) *100) %>%
  arrange(desc(percent))

max_reads_per_device <- max(reads_per_device$n)

avg_reads_per_device <- mean(reads_per_device$n)

device_count <- length(unique(reads_per_device$device))

reads_missing_address <- df %>%
  filter(missing_address %in% c('TRUE'))

devices_missing_address = reads_missing_address %>%
  filter(device== "SPD2020537MDC" | device == "BXH9904 SPD2020627MDC")  %>%
  nrow()
devices_missing_address_p = devices_missing_address / total_reads * 100

peo_device_reads = sum(df$device== "PEO508459" | df$device == "PEOATG508533") 
peo_device_reads_p = peo_device_reads / total_reads * 100


```

The data set consists of **`r device_count`** devices. Each device captured an average of **`r round(avg_reads_per_device, 0)`** license plates, with the most device capturing **`r max_reads_per_device`** plates. Two devices were completely missing addresses, comprising **`r round(devices_missing_address_p, 0)`%** of all reads in the data. Another two devices were attached to parking enforcement officer vehicles (PEO). Reads from PEO devices account for **`r round(peo_device_reads_p, 0)`%** of reads in the data set.

### Figure 6: Reads Per Device

```{r table of devices, echo=FALSE, message=FALSE}

table_reads_per_device <- kable(reads_per_device, 
                                col.names=c('Device', 'Reads',' Percent of Reads'),
                                row.names = F) %>%
     kable_styling(
       bootstrap_options = "bordered",
       position = "center",
       full_width = FALSE
     )

table_reads_per_device

```

### Figure 7: Tracking the Movements of ALPR Devices

The path of two active ALPR devices are depicted below. The maps show where the device captured license plates on camera througout the day. 

```{r map device path, echo=FALSE, message=FALSE, warning=FALSE}

# Create base of Seattle
base <- get_map("seattle, washington", zoom = 11, maptype = c("roadmap"))

# Filter data for specific device
device_M207541 <- df %>%
  arrange(device, dt) %>%
  filter(device == "M207541_Deleted_2021-10-21_22 50:42")

# Map device path 
date_label <- c(
  '2021-10-01'="Friday",
  '2021-10-02'="Saturday",
  '2021-10-03'="Sunday",
  '2021-10-04'="Monday",
  '2021-10-05'="Tuesday",
  '2021-10-06'="Wednesday",
  '2021-10-07'="Thursday",
  '2021-10-08'="Friday"
)

device_M207541_map <- ggmap(base) + 
  geom_path(data = device_M207541, aes(x = lon, y = lat, color = hour(dt),
                                       group = device)) + 
  facet_wrap(~date, strip.position = "bottom",
             labeller = as_labeller(date_label)) +
  theme_void() + 
   theme(
     plot.caption = element_text(face = "italic"), 
     legend.key.width = unit(30, "pt"), 
     legend.position	= "bottom", 
     legend.justification	= "center", 
     legend.box = "horizontal",
     legend.margin = margin(6, 6, 6, 6), 
     strip.background = element_rect(colour = "black", 
                                     fill = "white", size = .2),
     strip.text.x = element_text(face = "bold", size = 10, 
                                margin = margin(4, 2, 4, 2, "pt")),
     panel.spacing = unit(10, "pt")) + 
    labs(
         subtitle = "This vehicle surveilled all throughout Seattle across multiple days.",
         caption = "Source: SPD",
         color = "Time of Day")

device_M207541_map

```

```{r map device path to prison, echo=FALSE, message=FALSE, warning=FALSE}

everett <- get_map("everett, washington", zoom = 11, maptype = c("roadmap"))

device_SPD2020621MDC <- df %>%
  arrange(device, dt) %>%
  filter(device == "SPD2020621MDC")

device_SPD2020621MDC_map <- ggmap(everett) + 
  geom_path(data = device_SPD2020621MDC, aes(x = lon, y = lat, color = hour(dt),
                                       group = device)) + 
  facet_wrap(~date, strip.position = "bottom",
             labeller = as_labeller(date_label)) +
  theme_void() + 
   theme(
     plot.caption = element_text(face = "italic"), 
     legend.key.width = unit(30, "pt"), 
     legend.position	= "bottom", 
     legend.justification	= "center", 
     legend.box = "horizontal",
     legend.margin = margin(6, 6, 6, 6), 
     strip.background = element_rect(colour = "black", 
                                     fill = "white", size = .2),
     strip.text.x = element_text(face = "bold", size = 10, 
                                margin = margin(4, 2, 4, 2, "pt")),
     panel.spacing = unit(10, "pt")) + 
    labs(
      subtitle = "Meanwhile, this vehicle was active during just one day in the data set, and it only captured plates along the route to the local prison.",
      caption = "Source: SPD",
      color = "Time of Day")

device_SPD2020621MDC_map

```


## Examining ALPR System Usage During the Day

```{r describe date and time, echo=FALSE, message=FALSE}

reads_per_day <- df %>% 
  group_by(wday) %>% 
  summarize(n = n()) 
max_reads_per_day <- max(reads_per_day)

reads_per_min <- df %>% 
  group_by(day, hour, minute) %>% 
  summarize(n = n()) 
max_reads_per_min <- max(reads_per_min)

```

Contrary to our expectations, SPD devices were overall relatively inactive during the beginning of the work day and inactive during the morning rush hour commute. **Figure 8** and **Figure 9** show that devices collected the most data in the middle of the work day, between **12:00 pm** and **6:00 pm**. On the busiest day in our data set, SPD cameras captured **`r max_reads_per_day`** license plates. Up to **`r max_reads_per_min`** plates were captured in a single minute. 

### Figure 8: SPD ALPR Collection by Time of Day

ALPR activity starts low at the beginning of the day and continues to progress over time. Overall, ALPR activity was busiest in the middle of the day. 

```{r bar plot reads by hour, echo=FALSE, message=FALSE, warning=FALSE}

bar_reads_per_hour <- df %>%
  ggplot( aes(x=hour)) +
    geom_bar(bins = 24, binwidth=1, 
                    fill="#4b2e83", color="#4b2e83", alpha=0.9) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = ampm,
    breaks = function(z) pretty(z, n = 12)) +
         theme(
            axis.title = element_text(face = "bold"),
            axis.text.x = element_text(margin = margin(t = .3, unit = "cm")),
           plot.caption = element_text(face = "italic")) +
   labs(y = "Reads per Hour", 
        x = "Hour of Day",
        caption = "Source: SPD")

bar_reads_per_hour

``` 

### Figure 9: SPD ALPR Collection by Time of Day Per Day

This trend applies to each day of the week. 

```{r bar plot reads by hour per day, echo=FALSE, message=FALSE, warning=FALSE}

bar_reads_per_hour_per_day <- df %>%
  ggplot( aes(x=hour)) +
    geom_bar(bins = 24, binwidth=1, 
                    fill="#4b2e83", color="#4b2e83", alpha=0.9) +
    facet_wrap(~date,  
               ncol = 4,
               shrink = FALSE,
               strip.position = "bottom",
               labeller = as_labeller(date_label)) + 
  scale_y_continuous(labels = comma) +
  scale_x_continuous(labels = ampm,
                     breaks = function(z) pretty(z, n = 12)) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(margin = margin(t = .3, unit = "cm")), 
    strip.background = element_rect(colour = "black",fill = "white", size = .2),
    strip.text.x = element_text(face = "bold", size = 8, 
                                margin = margin(4, 2, 4, 2, "pt")),
    plot.caption = element_text(face = "italic")) +
   labs(y = "Reads per Hour", 
        x = "Day of the Week",
        caption = "Source: SPD")

bar_reads_per_hour_per_day

```

### Figure 10: SPD ALPR Collection by Day of the Week
ALPR activity was especially busier on weekdays, with a lower frequency of ALPR captures occurring over the weekend. 

```{r bar plot reads by day, echo=FALSE, message=FALSE, warning=FALSE}

bar_reads_per_day <- df %>%
  ggplot(aes(x=wday)) +
    geom_bar( bins = 7, binwidth=.5, 
                    fill="#4b2e83", color="#4b2e83", alpha=0.9) +
  scale_y_continuous(labels = comma) +
  scale_x_continuous(breaks = seq(1,7, by = 1), labels = c(
    "Sunday","Monday","Tuesday","Wednesday","Thursday", "Friday", "Saturday")) +
  theme(
            axis.title = element_text(face = "bold"),
            axis.text.x = element_text(margin = margin(t = .3, unit = "cm")),
           plot.caption = element_text(face = "italic")) +
  labs(y = "Reads Per Day", 
          x = "Day of the Week",
          caption = "Source: SPD") 

bar_reads_per_day

```

### Figure 10: SPD ALPR Collection by Day and Device

Not all ALPR devices recorded license plate reads every day of the week.

```{r bar plot reads by day and device, echo=FALSE, message=FALSE, warning=FALSE}

## WHAT'S MEANINGFUL TO LOOK AT: chart of missingness by device by date and time (figure 7b but add date)

#Proportions of the bar colored by devices
bar_reads_per_day_device <- df %>%
  ggplot(aes(x=wday)) +
  geom_bar(aes(fill = device),
           bins = 7, binwidth=.5, alpha=0.9) +
    scale_y_continuous(labels = comma) +
    scale_x_continuous(breaks = seq(1,7, by = 1), labels = c(
      "Sunday","Monday","Tuesday","Wednesday","Thursday", "Friday", "Saturday")) +
    scale_fill_manual("ALPR Devices", values = c(
      "BXH9904 SPD2020627MDC" = "gray", 
      "M207541_Deleted_2021-10-21_22 50:42" = "#571B7E", 
      "M207709" = "#663399", 
      "PEO508459" = "#800080", 
      "PEOATG508533" = "#967BB6",
      "SPD2020537MDC" = "gray", 
      "SPD2020602MDC" = "#2E1A47", 
      "SPD2020621MDC" = "#9D00FF", 
      "SPD2020681MDC" = "#4b2e83")) + 
  theme(
    legend.position = "none", 
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(margin = margin(t = .3, unit = "cm")),
    plot.caption = element_text(face = "italic")) +
     labs(y = "Reads Per Day", 
        x = "Day of the Week",
        caption = "Source: SPD")

bar_reads_per_day_device

```

